---
title: "Homework 6"
author: "Stephanie Solis (PID: A18572681)"
format: pdf
editor: visual
---

## CODE

```{r}
library(bio3d)
```

```{r}
analyze_drug_binding <- function(pdb_id) {

  pdb <- read.pdb(pdb_id)
  pdb_chainA <- trim.pdb(pdb, chain = "A", elety = "CA")
  b_factors <- pdb_chainA$atom$b

  plotb3(b_factors,
         sse = pdb_chainA,
         typ = "l",
         ylab = "B-factor",
         main = paste("B-factor plot for", pdb_id))

  return(b_factors)
}
```

```{r}
b1 <- analyze_drug_binding("4AKE")
b2 <- analyze_drug_binding("1AKE")
b3 <- analyze_drug_binding("1E4Y")
```

## QUESTIONS

> Q. Comments on what are the inputs to the function.

The input of the function is pdb_id, a character string that tells the function which protien structure to load and analyze.

> Q. what the function does and how to use it.

The function reads a protein structure from the PDB, extracts only chain A and the C‑alpha atoms, pulls out their B‑factors, and makes a line plot showing the B‑factor pattern for that protein. You use the function by giving it a PDB ID as a character string. 

> Q. what is the output of the function

The output of the function is a numeric vector of B‑factors for chain A C‑alpha atoms from the protein you selected.
