---
title: "Class12"
author: "Stephanie Solis (A18572681)"
format: pdf
editor: visual
---

```{r}
url <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"
dat <- read.table(url, header = TRUE, stringsAsFactors = TRUE)

```


```{r}
head(dat)
str(dat)

```

```{r}
table(dat$geno)
```

> Q13. 

The three genotypes show clear differences in ORMDL3 expression. The median expression for individuals with the A/A genotype is 31.24847, which is the highest among the groups. The A/G heterozygotes show an intermediate median expression of 25.06486, while the G/G genotype has the lowest median expression at 20.07363. This pattern suggests a stepwise decrease in ORMDL3 expression as the number of G alleles increases.

```{r}
tapply(dat$exp, dat$geno, median)

```



```{r}
boxplot(exp ~ geno, data = dat,
        xlab = "Genotype at rs8067378",
        ylab = "ORMDL3 Expression",
        main = "ORMDL3 Expression by Genotype",
        col = c("lightblue", "lightgreen", "lightpink"))

```
>Q14. 

The boxplot comparing A/A, A/G, and G/G genotypes shows a clear downward trend in ORMDL3 expression from A/A to G/G. Because individuals with the G/G genotype express ORMDL3 at substantially lower levels than those with A/A, the data indicate that rs8067378 does influence ORMDL3 expression. In other words, this SNP acts as an eQTL, with the G allele associated with reduced expression of the gene.
