---
title: "Class14"
author: "Stephanie Solis (PID A18572681)"
format: pdf
editor: visual
---

## Differential Expression Analysis

```{r}
library(DESeq2)
```
```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"
# Import metadata and take a peek
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```

> Q. Complete the code below to remove the troublesome first column from count-Data

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[, -1])
head(countData)
```


## Set up for DESeq

```{r}
dds <- DESeqDataSetFromMatrix(countData = countData,
colData = colData,
design = ~condition)
```

```{r}
rownames(colData) == colnames(countData)
```

## Remove Zero count

> Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns). Tip: What will rowSums() of countData return and how could you use it in this context?

```{r}
# Filter count data where you have 0 read count across all samples.
countData = countData[rowSums(countData) > 0, ]
head(countData)

```

## Running DESeq2

```{r}
dds = DESeq(dds)
```

```{r}
res = results(dds)
```

> Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
summary(res)
```

## Volcano plot

```{r}
library(ggplot2)
ggplot(res) + aes(log2FoldChange, -log(padj)) +
geom_point()
```


> Q. Improve this plot by completing the below code, which adds color, axis labels and cutoff lines:


```{r}
# Make a color vector for all genes
mycols <- rep("grey", nrow(res))
# Color blue the genes with fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "blue"
# Color gray those with adjusted p-value more than 0.01
mycols[ res$padj > 0.01 ] <- "grey"
4
ggplot(res) +
aes(log2FoldChange, -log(padj))+
geom_point(color = mycols) +
geom_vline(xintercept = c(-2,2)) +
geom_hline(yintercept = -log(0.01))
```
>Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below. 

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)

res$symbol <- mapIds(
  org.Hs.eg.db,
  keys = row.names(res),
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)

```

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
keys = row.names(res),
keytype = "ENSEMBL",
column = "ENTREZID",
multiVals = "first")
```

```{r}
res$name <- mapIds(org.Hs.eg.db,
keys = row.names(res),
keytype = "ENSEMBL",
column = "GENENAME",
multiVals = "first")
```

```{r}
head(res, 10)
```

> Q. Finally for this section let’s reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$pvalue),]
write.csv(res, file = "deseq_results.csv")
```

## Pathway Analysis

```{r}
library(pathview)
```

```{r}
library(gage)
```

```{r}
library(gageData)
data(kegg.sets.hs)
data(sigmet.idx.hs)
# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]
# Examine the first 3 pathways
head(kegg.sets.hs, 3)

```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```


```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```
```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]
# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```
> Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-regulated pathways?

```{r}
## Top 5 down‑regulated pathways
keggrespathways_down <- rownames(keggres$less)[1:5]

# Extract the 8‑character KEGG IDs
keggresids_down <- substr(keggrespathways_down, start = 1, stop = 8)
keggresids_down

```
```{r}
# Plot pathview figures for the top 5 down‑regulated pathways
pathview(gene.data = foldchanges,
         pathway.id = keggresids_down,
         species = "hsa")

```


## Gene Ontology (GO)

```{r}
data(go.sets.hs)
data(go.subs.hs)
# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]
gobpres = gage(foldchanges, gsets=gobpsets)
lapply(gobpres, head)
```


## Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

> Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The most significant pathway is the top pathway in the Reactome results with the smallest Entities p-value. It may differ from KEGG results because Reactome and KEGG use different pathway definitions and analysis methods.
