---
title: "Class 5: Data Viz with ggplot"
author: "Stephanie Solis (PID A18572681)"
format: pdf
toc: true
---

## Background
There are lot's of ways to amek figures in R. These include so-called "base R" graphics (e.g. `plot()`) and tones of add-on packages **ggplot2**.

For example here we make the same plot with both:
```{r}
head(cars)
```

First I need to install the package with the command `install.package()`

> **N.B.** We never run on install cmd in a quarto code chunk or we will end up re-installing packages many many times-which is not what we want!

Everytime we want to use of these "add-on" packages we need to load it up in R with the `lubrary()` function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data map to the plot
- The **geom**etry, the type of plot 

```{r}
p <- ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() + geom_smooth(method="lm",se=FALSE) + labs(title="Stopping distance of old cars", subtitle = "Data from the 'cars' object", x="Speed (mph)", y="Distamce (ft)")
```

render it out
```{r}
p
```

```{r}
p+theme_bw()
```

## Gene expression plot

We can read the input data from the class website 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot

```{r}
ggplot(genes)+ aes(Condition1, Condition2)+ geom_point()
```

```{r}
table(genes$State)
```

Version 2 let's color by `State` so we can see the "up" and "down" significant genes compared to all the unchanging genes

```{r}
ggplot(genes)+ aes(Condition1, Condition2, col=State)+ geom_point()
```

Version 3 plots, let's modify the default colors to something we like

```{r}
ggplot(genes)+ aes(Condition1, Condition2, col=State)+ geom_point() + scale_colour_manual( values=c("red", "blue", "magenta")) + labs(x="Control (no drugs)", y="Drug", title= "Gene Expressin Changes upon GLP-1 drug") + theme_bw()
```

## Going Further

Let's have a look at the famous **gapminder** dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(gapminder)+ aes(gdpPercap, lifeExp,  col=continent) + geom_point(alpha=0.3)
```

Let's "facet (i.e. make a separate plot) by continent rather than the big hot mess above. 

```{r}
ggplot(gapminder) + aes(gdpPercap, lifeExp, col=continent) + geom_point(alpha=0.3) + facet_wrap(~continent)
```

## Custom plots

How big is this gapminder dataset?

```{r}
nrow(gapminder)
```

I want to "filter" down to a subset of this data. I wil use the **dplyr** package to help me. 

First I need to install it and then load it up....
install.packages("dplyr") and then `library(dyplr)`

```{r}
library(dplyr)

gapminder_2007 <-filter(gapminder, year==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country=="Ireland")
```

```{r}
gapminder_2007 <-filter(gapminder, year==2007)
head(gapminder_2007)

filter(gapminder_2007, country=="United States")
```

> Q. Make a plot comparing 1997 and 2007 for all countries 

```{r}
input <- filter(gapminder, year %in% c(1977, 2007))
head (input)
```

```{r}
input <- filter(gapminder, year %in% c(1977,2007))

ggplot(data=input, aes(gdpPercap, lifeExp, col=continent)) + geom_point() +facet_wrap(~year)
```




